---
title: "gov51 pset2"
author: "Khalid Alzamil"
format: pdf
editor: visual
---

## 

#### **1 Part 1: The Original Study (20 points)**

**1.1 Exploring the Experiment**
1.1.1 Load packages and data

```{r}
library(tidyverse)
gay <- read_csv("data/raw/gay.csv")
```

Q1.1 (3 points): How many total observations are in gay.csv? How many unique values does
study take? List all the treatment conditions in the treatment variable.

```{r}
gay <- read_csv("data/raw/gay.csv")

# number of observations
nrow(gay)

# number of unique studies
n_distinct(gay$study)

# list treatment conditions
unique(gay$treatment)
```

*The dataset contains 69,592 observations.*

*The variable study takes 2 unique values.*

*The treatment conditions are: No Contact, Recycling Script by Gay Canvasser, Same sex Marriage Script by Gay Canvasser, Recycling Script by Straight Canvasser, and Same sex Marriage Script by Straight Canvasser.*

Q1.2 (3 points): Filter to Study 1, Wave 1 only. Calculate the mean of ssm (same-sex marriage attitude, scored 1–5) for each treatment group using group_by() and summarize(). Which
treatment group shows the highest average support?

```{r}
gay %>%
  filter(study == 1, wave == 1) %>%
  group_by(treatment) %>%
  summarize(mean_ssm = mean(ssm, na.rm = TRUE)) %>%
  arrange(desc(mean_ssm))
```

*The Recycling Script by Gay Canvasser treatment shows the highest average support for same-sex marriage*

Q1.3 (3 points): Still using Study 1, Wave 1, calculate the standard deviation of ssm within
each treatment group. Are the standard deviations roughly similar across groups? Why is this
important in a randomized experiment? (2–3 sentences)

```{r}
gay %>%
  filter(study == 1, wave == 1) %>%
  group_by(treatment) %>%
  summarize(sd_ssm = sd(ssm, na.rm = TRUE))
```

Your interpretation:
The standard deviations are very similar across the treatment groups, all around 1.67–1.70. This suggests that the variability in same sex marriage attitudes is roughly the same across groups. In a randomized experiment similar variability across treatment groups is important because it indicates that the groups are comparable + that differences in outcomes are more likely due to the treatment rather than underlying differences between groups.

Q1.4 (3 points): Create a bar plot of the mean ssm by treatment group (Study 1, Wave 1) using
ggplot() with geom_col(). Hint: First create a summary data frame with the means, then plot
it.

```{r}
mean_ssm <- gay %>%
  filter(study == 1, wave == 1) %>%
  group_by(treatment) %>%
  summarize(mean_ssm = mean(ssm, na.rm = TRUE))

ggplot(mean_ssm, aes(x = treatment, y = mean_ssm)) +
  geom_col() +
  labs(
    title = "Average Same-Sex Marriage Support by Treatment",
    x = "Treatment Group",
    y = "Mean SSM"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) #names were too long
```

**1.2 Checking Balance and Comparing Outcomes**
Q1.5 (4 points): A well-run randomized experiment should produce groups that look similar at
baseline. Filter gay.csv to Study 1, Wave 1. Compare the “Same-Sex Marriage Script by Gay
Canvasser” group and the “No Contact” group:
• Calculate the mean and SD of ssm for each group
• Compute the standardized difference in means:
A standardized difference below 0.25 in absolute value suggests good balance. Do these groups look
similar at baseline? (2–3 sentences)

```{r}
baseline <- gay %>%
  filter(
    study == 1,
    wave == 1,
    treatment %in% c(
      "Same-Sex Marriage Script by Gay Canvasser",
      "No Contact"
    )
  )

baseline %>%
  group_by(treatment) %>%
  summarize(
    mean_ssm = mean(ssm, na.rm = TRUE),
    sd_ssm = sd(ssm, na.rm = TRUE)
  )
```

```{r}
means <- baseline %>%
  group_by(treatment) %>%
  summarize(
    mean_ssm = mean(ssm, na.rm = TRUE),
    sd_ssm = sd(ssm, na.rm = TRUE)
  )

m1 <- means$mean_ssm[means$treatment == "Same-Sex Marriage Script by Gay Canvasser"]
m2 <- means$mean_ssm[means$treatment == "No Contact"]

s1 <- means$sd_ssm[means$treatment == "Same-Sex Marriage Script by Gay Canvasser"]
s2 <- means$sd_ssm[means$treatment == "No Contact"]

std_diff <- (m1 - m2) / sqrt((s1^2 + s2^2)/2)

std_diff
```

Your interpretation: The mean and standard deviation are very similar between the Same sex Marriage Script by Gay Canvasser group and the No Contact group. The standardized difference in means is approx -0.01, aka below the 0.25 threshold. This suggests that the groups are well balanced at baseline which is what we would expect from a properly randomized experiment.

Q1.6 (4 points): Now look at the outcomes. Filter to Study 1, Wave 2 only. Calculate the mean
ssm for the gay canvasser group and the No Contact group. What is the difference in means (gay
canvasser minus no contact)?
Interpret: since the groups were balanced at baseline, what does a difference at Wave 2 suggest?
(2–3 sentences)

```{r}
wave2 <- gay %>%
  filter(
    study == 1,
    wave == 2,
    treatment %in% c(
      "Same-Sex Marriage Script by Gay Canvasser",
      "No Contact"
    )
  )

wave2_means <- wave2 %>%
  group_by(treatment) %>%
  summarize(mean_ssm = mean(ssm, na.rm = TRUE))

wave2_means

#diff in means
diff <- wave2_means$mean_ssm[wave2_means$treatment == "Same-Sex Marriage Script by Gay Canvasser"] -
        wave2_means$mean_ssm[wave2_means$treatment == "No Contact"]

diff
```

Your interpretation: *At Wave 2, the mean support for same sex marriage in the “Same sex Marriage Script by Gay Canvasser” group is about 0.10 points higher than in the “No Contact” group. Because the groups were balanced at baseline this difference suggests that the canvassing treatment may have caused an increase in support for same sex marriage which indicates a possible causal effect of the treatment on attitudes toward same sex marriage.*

#### 2 Part 2: The Forensic Discovery (30 points)

**2.1 The Baseline Distribution**

Q2.1 (3 points): Load gayreshaped.csv and ccap2012.csv. Report the number of observations
and column names for each dataset. In gayreshaped.csv, the variables therm1 through therm4
represent feeling thermometer scores (0–100) toward gay men and lesbians, measured at each survey
wave.

```{r}
gay_r <- read_csv("data/raw/gayreshaped.csv")
ccap <- read_csv("data/raw/ccap2012.csv")

#observations
nrow(gay_r)
nrow(ccap)

#names
colnames(gay_r)
colnames(ccap)
```

The dataset gay_r contains 11948 observations with the variables: study, treatment, therm1, therm2, therm3, and therm4. The dataset ccap contains 43998 observations with variables: caseid and gaytherm.

Q2.2 (5 points): Extract the baseline feeling thermometer (therm1) for Study 1 from
gayreshaped.csv. Calculate its mean, median, and standard deviation. Then do the same for
gaytherm from ccap2012.csv (remove NAs). Report both sets of statistics side by side. What do
you notice? (1–2 sentences)

```{r}
# baseline thermometer
gay_base <- gay_r %>%
  filter(study == 1) %>%
  summarize(
    mean = mean(therm1, na.rm = TRUE),
    median = median(therm1, na.rm = TRUE),
    sd = sd(therm1, na.rm = TRUE)
  )

# ccap thermometer
ccap_base <- ccap %>%
  summarize(
    mean = mean(gaytherm, na.rm = TRUE),
    median = median(gaytherm, na.rm = TRUE),
    sd = sd(gaytherm, na.rm = TRUE)
  )

# report
baseline_stats <- bind_rows(
  gayreshaped = gay_base,
  ccap2012 = ccap_base,
  .id = "dataset"
)

baseline_stats
```

What do you notice?

The two datasets have a very similar average thermometer scores toward gay men and lesbians. The means are almost identical (about 58), and the standard deviations are also close, suggesting that the overall level and variation of attitudes toward gay men and lesbians are quite similar in the experimental sample and the CCAP survey.

Q2.3 (5 points): Create side-by-side histograms comparing the LaCour Study 1 baseline (therm1)
and the CCAP data (gaytherm). Use the same bin breaks for both: breaks = seq(0, 100, by
= 5). Hint: Combine the two datasets into one data frame with a column indicating the source, then use
ggplot() with facet_wrap(). Do these look like they came from different populations? (2–3 sentences)

```{r}
# extract thermometers
gay_base <- gay_r %>%
  filter(study == 1) %>%
  select(therm = therm1) %>%
  mutate(source = "LaCour Study 1")

ccap_base <- ccap %>%
  select(therm = gaytherm) %>%
  mutate(source = "CCAP")

combined <- bind_rows(gay_base, ccap_base) %>%
  filter(!is.na(therm))

# histogram 
ggplot(combined, aes(x = therm)) +
  geom_histogram(breaks = seq(0, 100, by = 5), color = "black", fill = "steelblue") +
  facet_wrap(~source) +
  labs(
    title = "Distribution of Feeling Thermometer Toward Gay Men and Lesbians",
    x = "Feeling Thermometer (0–100)",
    y = "Count"
  )
```

Your interpretation: *Although they have very different numbers of observation, the two histograms have very similar shapes and cover the same range of thermometer scores. Both distributions show clustering around the middle and higher values, suggesting similar attitudes toward gay men and lesbians, which suggests the LaCour Study 1 baseline sample doesn't look dramatically different from the CCAP population.*

**2.2 Re-test Reliability**
When real people take the same survey twice, their answers correlate — but not perfectly. People’s
moods change, they round differently, they misread questions. A re-test correlation of 0.95–0.97 is
normal for feeling thermometers. A correlation near 1.0 is suspicious.

Q2.4 (6 points): For both the Study 1 and Study 2 control groups (“No Contact”), calculate:
(a) The correlation between therm1 and therm2 (use cor() with use = "complete.obs")
(b) The standard deviation of the within-person change (therm2 - therm1)
Present your results side by side. A re-test correlation of 0.95–0.97 is typical for real feeling
thermometer data. Which study shows a more plausible pattern, and why? (3–4 sentences)

```{r}
control_stats <- gay_r %>%
  filter(treatment == "No Contact") %>%
  mutate(change = therm2 - therm1) %>%
  group_by(study) %>%
  summarize(
    correlation = cor(therm1, therm2, use = "complete.obs"),
    sd_change = sd(change, na.rm = TRUE)
  )

control_stats
```

Your interpretation: *Study 2 shows a more plausible pattern of responses. Its correlation between therm1 and therm2 (0.973) falls within the typical range for retest reliability of feeling thermometers (0.95–0.97), and the standard deviation of the change is relatively larger, indicating meaningful variation in responses over time. Study 1 shows a very high correlation (0.998) and a very small standard deviation of change, suggesting that respondents’ answers barely changed between waves and this level of consistency is unusually high for actual survey data.*

Q2.5 (5 points): Create two scatter plots side by side:
(a) Study 1 control group: therm1 (x-axis) vs therm2 (y-axis), with a 45-degree line
(b) Study 2 control group: therm1 (x-axis) vs therm2 (y-axis), with a 45-degree line
Describe the differences. Which looks more like real longitudinal data, and why? (3–4 sentences)

```{r}
control_data <- gay_r %>%
  filter(treatment == "No Contact")

ggplot(control_data, aes(x = therm1, y = therm2)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  facet_wrap(~study) +
  labs(
    title = "Re-test Reliability of Feeling Thermometer",
    x = "Thermometer Wave 1 (therm1)",
    y = "Thermometer Wave 2 (therm2)"
  )
```

Your interpretation: *Study 1shows points almost perfectly aligned along the line showing that respondents answers changed very little between waves. In contrast, the Study 2 plot shows much more dispersion around the diagonal, with respondents’ answers increasing or decreasing between waves. This level of variation is typical for real survey data because responses fluctuate due to measurement error or changing attitudes. As such, Study 2 appears more consistent with realistic longitudinal survey data.*

**2.3 Putting It All Together**
Q2.6 (6 points): You’ve now found three irregularities:
1. The baseline distributions match an existing national survey
2. The re-test correlations are implausibly high (Study 1)
3. The within-person changes are implausibly small (Study 1)
In 4–5 sentences, explain why these three findings together suggest the data were fabricated rather
than collected in a real experiment. Use specific numbers from your analysis. Why would a
fabricator produce data with these properties?
Your interpretation:

The baseline distribution of feeling thermometer scores in the LaCour data closely matches the CCAP survery with nearly identical means (58.38 in the experiment vs. 58.71 in CCAP) and very similar standard deviations (28.50 vs. 29.37). However, the retest correlation in Study 1 is very high (0.998), far above the typical range of 0.95–0.97 for real feeling thermometer data and the standard deviation of the within-person change in Study 1 is only about 1.99, compared to 6.61 in Study 2, indicating that responses in Study 1 barely changed between waves. Taken together, these patterns suggest the data were likely fabricated rather than collected from real respondents. A fabricator might copy the distribution of a real survey like CCAP to make the data look realistic but unintentionally generate responses that are too consistent across waves which produces impossibly high correlations and minimal individual change.
